---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit README.Rmd, then
     render with `Rscript -e 'rmarkdown::render("README.Rmd")'`.
     Chunks that invoke `ry` run only when the binary is on PATH
     (docs CI builds it first), so the output below is real, not
     hand-written. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  error    = TRUE   # `ry check` exits non-zero on findings; that is the point
)
has_ry <- nzchar(Sys.which("ry"))
knitr::opts_chunk$set(eval = has_ry)
```

# ry

[![CI](https://github.com/sims1253/ry/actions/workflows/ci.yml/badge.svg)](https://github.com/sims1253/ry/actions/workflows/ci.yml)

`ry` is a fast static checker for the R language, written in Rust and
inspired by [astral-sh/ty](https://github.com/astral-sh/ty). It parses R
source with [tree-sitter-r](https://github.com/r-lib/tree-sitter-r),
infers types across your whole project, and reports likely bugs before
you run the code: calling a non-function, arithmetic on incompatible
types, misspelled data frame columns, unbound variables, malformed `if`
conditions, and more.

R will happily coerce, recycle, and partially match its way past most of
these mistakes at runtime and hand you a statistically wrong answer
instead of an error. ry's job is to catch the mistake while it is still
cheap.

## Who this is for

ry is built to support **education, practice, and research around
principled Bayesian workflows** in R:

- **Education.** Every diagnostic explains what R would silently do at
  runtime, so the checker doubles as a teacher of R's semantics. The
  [rule reference](docs/rules/) is written as teaching material, and the
  repository ships graded [exercises](exercises/) and a fully worked
  [Bayesian workflow demo](demos/bayesian-workflow/) -- plus a
  [broken twin](demos/bayesian-workflow-broken/) with seeded, realistic
  workflow bugs for practice.
- **Practice.** The demo workflow uses the modern parallel idiom --
  `purrr::in_parallel()` on `mirai` daemons -- and ry type-checks the
  parallel code exactly like its sequential equivalent.
- **Research.** Diagnostics are available as structured JSON
  (`--output-format json`, `--statistics`), so ry can instrument code
  quality across a corpus or a classroom. See
  [docs/architecture](docs/architecture.qmd) for the methodology,
  including how ry's claims are validated against R itself.

ry is **not** a formatter (pair it with `air` or `styler`) and not a
replacement for `lintr`'s style rules. It focuses on type- and
scope-driven diagnostics that need a whole-program view, and it prefers
staying silent over crying wolf: every rule is validated against an R
oracle and real CRAN code so that a warning from ry is worth reading.

## Install

ry is a Cargo workspace; build from source (Rust 1.82 or newer):

```sh
git clone https://github.com/sims1253/ry
cd ry
cargo build --release
# binary at target/release/ry
```

Prebuilt binaries and an R-side installer are planned; see `PLAN.md`.

## Quickstart

Point `ry check` at files, directories, or a project root. `.R`, `.Rmd`,
and `.qmd` files are picked up; R chunks in Quarto and R Markdown
documents are checked with diagnostics mapped back to the host file's
line numbers.

```{bash quickstart}
ry check examples/smoke.R
```

Diagnostics use the `full` format by default: the offending line, a
marker under the exact span, and -- where ry can help -- a `help:` line
with a suggestion:

```
analysis.R:12:8: error: [RY060] data frame has no column `dispp`
    mean(d$dispp)
           ^~~~~
    help: did you mean `disp`? available columns: mpg, cyl, disp, ...
```

Exit codes are CI-friendly: non-zero when any error-level diagnostic
fires (`--exit-zero` overrides; `--error-on-warning` promotes warnings).

## Package awareness

ry tracks `library()` / `require()` calls and resolves functions against
per-package type stubs, so the same name means the right thing in
context: `filter()` is `stats::filter` until dplyr is loaded, and
`dplyr::filter(df, x > 0)` resolves the column `x` against `df`'s schema
either way. Stubs currently ship for base R (plus stats/utils), dplyr,
purrr, mirai, and a minimal Bayesian stack (posterior, brms, loo,
cmdstanr, bayesplot). Packages loaded outside the checked sources (e.g.
in a startup file) can be declared in `ry.toml`.

Parallel purrr code checks like sequential code:

```r
library(purrr)
library(mirai)

daemons(4)
fits <- map(simulations, in_parallel(function(s) fit_one(s)))
rhats <- map_dbl(fits, function(f) max_rhat(f))
```

`in_parallel()` is type-transparent to ry, and the `map_*` family checks
the callback's return mode against the requested vector type -- a
`map_dbl` whose callback returns character is a diagnostic, not a
runtime surprise halfway through your simulation study.

## Learning materials

| Where | What |
|---|---|
| `docs/` | Quarto site: getting started, the type model as a tour of R's runtime semantics, rule reference, editor setup, CI recipes, architecture. |
| `demos/bayesian-workflow/` | A principled Bayesian workflow (prior predictive checks, fitting, diagnostics, posterior predictive checks, parallel simulation-based calibration) that checks clean. |
| `demos/bayesian-workflow-broken/` | The same workflow with seeded bugs, one diagnostic each; `ANSWERS.md` explains the statistical consequence of every bug. |
| `exercises/` | Koan-style exercises: edit each file until `ry check` is green. Ordered to teach the type model incrementally. |

## Configuration (`ry.toml`)

Discovered by walking up from the checked path. All keys optional:

```toml
# Promote / demote / disable rules by code (RY040), name
# (invalid-arithmetic), or "all".
error  = ["RY040"]
warn   = ["RY070"]
ignore = ["RY033"]

# Packages attached outside the checked sources.
packages = ["dplyr"]

error-on-warning = false
exit-zero        = false
output-format    = "full"     # full | concise | json | github | gitlab | junit

# gitignore-style patterns, relative to this ry.toml's directory.
exclude = ["renv", "tests/snaps/**"]
```

CLI flags override the config only when passed explicitly.

## Inline suppression

```r
x <- bad  # ry: ignore                 # suppress all rules on this line
x <- bad  # ry: ignore[RY010, RY040]   # suppress specific rules
x <- bad  # noqa: RY010                # flake8/ruff-compatible alias

# ry: ignore                           # standalone: suppresses the next line
# ry: ignore-file                      # file-level, anywhere in the file
```

## Editors

`ry server` speaks the Language Server Protocol over stdio: diagnostics
as you type (debounced, cached parses), hover with inferred types,
go-to-definition, references, rename, completion, signature help, inlay
hints, folding, and quick-fix actions that insert suppression comments.
Setup guides for VS Code, Positron, Neovim, and Helix are in
[docs/editors](docs/editors.qmd).

## CI

`--output-format github` emits workflow-command annotations; `gitlab`
and `junit` cover the other major CI systems. `--statistics` prints
per-rule counts for corpus work. A minimal GitHub Actions step:

```yaml
- run: ry check --output-format github .
```

## Rules

The table below is generated from the checker's own rule registry at
render time, so it cannot drift from the binary. Defaults can be
overridden per-project; `ry rule RY040` prints the long-form
explanation with examples.

```{r rules-table, echo = FALSE, results = "asis"}
rules <- jsonlite::fromJSON(
  system2("ry", c("explain-rule", "--output-format", "json"), stdout = TRUE)
)
knitr::kable(rules[, c("code", "name", "severity", "summary")])
```

## How ry stays honest

- **The R oracle.** Every behavioral claim is checked against R itself:
  a fixture corpus is executed by an R driver (in parallel, via
  `purrr::in_parallel()` on mirai daemons -- the same idiom the demos
  teach) and ry must agree with R's verdict. Runs in CI;
  `cargo test -p ry-checker --test oracle -- --ignored` locally.
- **Vendored CRAN code.** ry runs over vendored real-world packages and
  snapshots every diagnostic; the snapshot is triaged line by line, and
  a rule that dominates the snapshot is treated as broken.
- **Self-verifying docs.** Each rule's documented bad example is run
  against the checker in the test suite; this README is rendered with
  the real binary.

Known gaps are tracked in `PLAN.md` (no S4/R6/environment modeling, no
NAMESPACE resolution, no NA tracking yet).

## Contributing and architecture

Start with [ARCHITECTURE.md](ARCHITECTURE.md) (crate map, the three-pass
project check, where to add a rule) and
[CONTRIBUTING.md](CONTRIBUTING.md) (build gate, fixture conventions, and
the no-false-positives bar every new rule must clear).

## License

MIT OR Apache-2.0.
