---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit README.Rmd, then
     render with the release binary on PATH:
       PATH="$PWD/target/release:$PATH" Rscript -e 'rmarkdown::render("README.Rmd")'
     Chunks that invoke `ry` run only when the binary is available, so
     the output below is real, not hand-written. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  error    = TRUE   # `ry check` exits non-zero on findings; that is the point
)
has_ry <- nzchar(Sys.which("ry"))
knitr::opts_chunk$set(eval = has_ry)
```

# ry

[![CI](https://github.com/sims1253/ry/actions/workflows/ci.yml/badge.svg)](https://github.com/sims1253/ry/actions/workflows/ci.yml)

`ry` is a fast static checker for the R language, written in Rust and
inspired by [astral-sh/ty](https://github.com/astral-sh/ty). It parses R
source with [tree-sitter-r](https://github.com/r-lib/tree-sitter-r),
infers types across your whole project, and reports likely bugs before
you run the code: calling a non-function, arithmetic on incompatible
types, misspelled data frame columns, unbound variables, malformed `if`
conditions, and more.

R will happily coerce, recycle, and partially match its way past most of
these mistakes at runtime and hand you a statistically wrong answer
instead of an error. ry's job is to catch the mistake while it is still
cheap.

## Who this is for

ry is built to support **education, practice, and research around
principled Bayesian workflows** in R:

- **Education.** Every diagnostic explains what R would silently do at
  runtime, so the checker doubles as a teacher of R's semantics. A
  docs site, worked Bayesian-workflow demos, and graded exercises are
  in progress -- see [ROADMAP.md](ROADMAP.md).
- **Practice.** ry understands the modern parallel idiom --
  `purrr::in_parallel()` on `mirai` daemons -- and type-checks parallel
  code exactly like its sequential equivalent.
- **Research.** Diagnostics are available as structured JSON
  (`--output-format json`) with per-rule counts (`--statistics`), so ry
  can instrument code quality across a corpus or a classroom. How ry
  validates its own claims against R is described in
  [ARCHITECTURE.md](ARCHITECTURE.md).

ry is **not** a formatter (pair it with `air` or `styler`) and not a
replacement for `lintr`'s style rules. It focuses on type- and
scope-driven diagnostics that need a whole-program view, and it prefers
staying silent over crying wolf: every rule is validated against an R
oracle and real CRAN code so that a warning from ry is worth reading.

## Install

ry is a Cargo workspace; build from source (Rust 1.82 or newer):

```sh
git clone https://github.com/sims1253/ry
cd ry
cargo build --release
# binary at target/release/ry
```

Prebuilt binaries and an R-side installer are on the
[roadmap](ROADMAP.md).

## Quickstart

Point `ry check` at files, directories, or a project root (`.R` and
`.r` files are collected recursively; Quarto/R Markdown chunk checking
is on the roadmap):

```{bash quickstart}
ry check examples/smoke.R
```

Diagnostics use the `full` format by default -- the offending line with
the span underlined -- and messages carry the context you need to act,
e.g. a column miss lists what IS there:

```{bash column-miss, echo = FALSE, include = FALSE}
mkdir -p /tmp/ry-readme && printf 'd <- data.frame(mpg = 1, disp = 2)\nm <- mean(d$dispp)\n' > /tmp/ry-readme/analysis.R
```

```{bash column-miss-run}
ry check /tmp/ry-readme/analysis.R
```

Exit codes are CI-friendly: non-zero when any error-level diagnostic
fires (`--exit-zero` overrides; `--error-on-warning` promotes warnings).

## Package awareness

ry tracks `library()` / `require()` calls and resolves functions against
per-package type stubs, so the same name means the right thing in
context: `filter()` is `stats::filter` until dplyr is loaded, and
`dplyr::filter(df, x > 0)` resolves the column `x` against `df`'s schema
either way. Stubs currently ship for base R (plus stats and utils),
dplyr, purrr, mirai, and a minimal Bayesian stack (brms, posterior,
loo, bayesplot, cmdstanr). Packages attached outside the checked
sources can be declared in `ry.toml`.

Parallel purrr code checks like sequential code, and the typed map
family is checked against its callback:

```{bash purrr, echo = FALSE, include = FALSE}
printf 'library(purrr)\nlibrary(mirai)\ndaemons(2)\nok  <- map_dbl(1:4, in_parallel(function(i) i * 2))\nbad <- map_dbl(1:4, function(i) as.character(i))\n' > /tmp/ry-readme/parallel.R
```

```{bash purrr-run}
ry check /tmp/ry-readme/parallel.R
```

`in_parallel()` is type-transparent to ry: a `map_dbl` whose callback
returns character is a diagnostic before the run, not a surprise
halfway through your simulation study.

## Configuration (`ry.toml`)

Discovered by walking up from the checked path. All keys optional:

```toml
# Promote / demote / disable rules by code (RY040), name
# (invalid-arithmetic), or "all".
error  = ["RY040"]
warn   = ["RY070"]
ignore = ["RY033"]

# Packages attached outside the checked sources.
packages = ["dplyr"]

error-on-warning = false
exit-zero        = false
output-format    = "full"     # full | concise | json | github | gitlab | junit

# gitignore-style patterns, relative to this ry.toml's directory.
exclude = ["renv", "tests/snaps/**"]
```

CLI flags override the config only when passed explicitly.

## Inline suppression

```r
x <- bad  # ry: ignore                 # suppress all rules on this line
x <- bad  # ry: ignore[RY010, RY040]   # suppress specific rules
x <- bad  # noqa: RY010                # flake8/ruff-compatible alias

# ry: ignore                           # standalone: suppresses the next line
# ry: ignore-file                      # file-level, anywhere in the file
```

## Editors

`ry server` speaks the Language Server Protocol over stdio: diagnostics
as you type (debounced, cached parses), hover with inferred types,
go-to-definition, references, rename, completion, signature help, inlay
hints, folding, and quick-fix actions that insert suppression comments.
Connect it from any LSP-aware editor (VS Code, Positron, Neovim,
Helix, ...).

## CI

`--output-format github` emits workflow-command annotations; `gitlab`
and `junit` cover the other major CI systems. `--statistics` prints
per-rule counts after a run -- useful for corpus work. A minimal GitHub
Actions step:

```yaml
- run: ry check --output-format github .
```

## Rules

The table below is generated from the checker's own rule registry at
render time, so it cannot drift from the binary. Defaults can be
overridden per-project; `ry rule RY040` prints the explanation for one
rule.

```{r rules-table, echo = FALSE, results = "asis"}
rules <- jsonlite::fromJSON(
  system2("ry", c("explain-rule", "--output-format", "json"), stdout = TRUE)
)
knitr::kable(rules[, c("code", "name", "severity", "summary")])
```

## How ry stays honest

- **The R oracle.** Every behavioral claim is checked against R itself:
  the fixture corpus is executed by an R driver -- in parallel, via
  `purrr::in_parallel()` on mirai daemons, the same idiom the checker
  models -- and ry must agree with R's verdict. Runs in CI;
  `cargo test -p ry-checker --test oracle -- --ignored` locally.
- **Vendored CRAN code.** ry runs over vendored real-world packages
  (currently glue and purrr) and snapshots every diagnostic; each
  snapshot line is triaged, and a rule that dominates a snapshot is
  treated as broken. The glue snapshot is empty.
- **A typeshed that cannot invent functions.** Every name in the
  embedded type stubs is verified against R in CI
  (`scripts/audit_typeshed.R`).

Known gaps: no S4 / R6 / environment modeling, no NAMESPACE resolution
(cross-package names outside the shipped stubs resolve to opaque), no
NA tracking yet. See [ROADMAP.md](ROADMAP.md).

## Contributing and architecture

Start with [ARCHITECTURE.md](ARCHITECTURE.md) (crate map, the three-pass
project check, where to add a rule) and
[CONTRIBUTING.md](CONTRIBUTING.md) (build gate, fixture conventions, and
the false-positive bar every new rule must clear).

## License

MIT OR Apache-2.0.
